Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-11 paper/main.tex
surgery to trim fat on Derivation System

A clean theoretical framework for compositional bounds sometimes
even allow to get better asymptotic bounds.  During our experiments
we found an example of this situation in some code from the cBench
benchmark suite.
%
\begin{lstlisting}
for (;;) {
    do { l++; tick(1); } while (l < h && nondet());
    do { h--; tick(1); } while (h > l && nondet());
    if (h <= l) break;
    tick(1); /* swap two elements */ }
\end{lstlisting}
%
The above snippet is the inner loop of a Quicksort implementation, more
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding into our analysis gives the
correct and tight worst-case bound $2+3|[l,h]|$.  With the amortized
analysis approach in mind, it is clear that the three loops get their
potential from the same interval $[l,h]$.  However, KoAT fails to
prove the termination of the loop and LOOPUS derives the quadratic bound
$(h-l-1)^2$.  Following the classic technique, these tools try to find one
linear ranking-function for each loop and combine them multiplicatively
or additively, resulting in a loose quadratic bound or no bound at all.

@@

\ifshort{
\begin{figure*}
\small
\begin{mathpar}
\Rule[leftskip=1.2cm,rightskip=.5cm]{Q:Ret}
{ P = Q_R[\Vret/x]
\\ \forall i \in \dom{P} . \, p_i = q_i
}
{ B; (\Gamma_R,Q_R); (\Gamma_R[\Vret/x], Q) \vdash \code{return}~x \dashv (\Gamma',Q') }
%
\and \Rule[leftskip=.5cm,rightskip=1.2cm]{Q:Update}
{ q'_{xy}, q'_{yx} \in \Qplusz
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
%
\\ \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\and \Rule[leftskip=.2cm,rightskip=.1cm]{Q:IncP}
{ \Gamma \models y \ge 0 \!
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx} \!
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule[leftskip=1cm,rightskip=.5cm]{Q:DecP}
{ \Gamma \models y \ge 0
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
}
%
\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
{ M = M_u + M_e(x {\pm} y)
\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:If}
{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
}
{ B; R; (\Gamma, Q{+}M_e(e)) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
%
\and \Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and\Rule{Q:Call}
{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f) \!
\\ \text{Loc} = \text{Locals}(Q) \!
\\ \forall i \neq j . \, x_i \neq x_j \!
\\ c \in \Qplusz \!
\\ Q = P + S \!
\\ Q' = P' + S
\\ U = Q_f[\Vargs / \vec x]
\\ U' = Q'_f[\Vret/r]
\\ \forall i \in \dom{U} . \, p_i = u_i
\\ \forall i \in \dom{U'} .\, p'_i = u'_i
\\ \forall i \not\in \dom{U'} .\, p'_i = 0
\\ \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
  \vdash r \gets f(\vec x) \dashv
  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
}
%
\and
\Rule[leftskip=.2cm,rightskip=.2cm]{Q:Extend}
{ \fenv f = (\vec y, S_f)
\\\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S_f \dashv (\Gamma', Q')
}
{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
%
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\ \!\! \Gamma_1 \models \Gamma_2
\\\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ \Gamma'_2 \models \Gamma'_1
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
}
{ Q' \succeq_\Gamma Q }

@@

Most interesting are the rules {\sc Q:IncP}, {\sc Q:DecP}, and {\sc Q:Inc}
which describe how the potential is distributed after a size change of
a variable.  The rule {\sc Q:IncP} is for increments $x \gets x + y$
and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
that can be applied if $y$ is not positive.  The rules are equivalent
in the case where $y=0$.  Finally, the rule {\sc Q:Inc} can be applied
if we cannot deduce any information about $y$.  In the implementation,
the rules for increments and decrements are combined into one syntax
directed rule where the specialized rules take priority over {\sc Q:Inc}.

To explain how rules for increment and decrement work, it is
sufficient to understand the rule {\sc Q:IncP}.  The others follow the
same idea and are symmetrical.  In {\sc Q:IncP}, the program updates a
variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the change of the
program state.  We write $x'$ for the value of $x$ after the assignment.
Since $x$ is the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized.  Note that for any $u$,
$\inter x u$ will get smaller with the update, and if $x' \in \inter x
u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
get bigger with the update.  We know that $|\inter v {x'}| \le y +
|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
$q_{vx}$ to pay for this change.  The rule ensures this
only for $v \not\in \mathcal U$ because we know that $x \le v$
otherwise, and thus $|\inter v x| = 0$.

Another interesting rule is {\sc Q:Call}.  It needs to account for the
changes to the stack caused by the function call, the arguments/return
value passing, and the preservation of local variables.  We can sum up
the main ideas of the rule as follows.
\itemskip
\begin{itemize}
\item The potential in the pre- and postcondition of the function
  specification is equalized to its matching potential in the callee's
  pre- and postcondition.
\itemskipIn
\item
  $\!$The potential of intervals $|\inter x y|$ is preserved
  if $x$ and $y$ are local\iffull{ variables.}{.}
\item
  The unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are set to zero in the postcondition.
\end{itemize}
\itemskip
If $x$ and $y$ are local variables and $f(x,y)$ is
called, {\sc Q:Call} splits the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P {+} S$ and $Q' =
P' {+} S'$.  Arguments in the function precondition
$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function's postcondition
$(\Gamma'_f, Q'_f)$.

>>>>

A clean theoretical framework for compositional bounds sometimes
even allow to get better asymptotic bounds.  During our experiments
we found an example of this situation in some code from the cBench
benchmark suite.
%
\begin{lstlisting}
for (;;) {
    do { l++; tick(1); } while (l < h && *);
    do { h--; tick(1); } while (h > l && *);
    if (h <= l) break;
    tick(1); /* swap two elements */ }
\end{lstlisting}
%
The above snippet is the inner loop of a Quicksort implementation, more
precisely, it is the partitioning part of the algorithm.  This partition
loop has linear complexity, and feeding into our analysis gives the
correct and tight worst-case bound $2+3|[l,h]|$.  With the amortized
analysis approach in mind, it is clear that the three loops get their
potential from the same interval $[l,h]$.  However, KoAT fails to
prove the termination of the loop and LOOPUS derives the quadratic bound
$(h-l-1)^2$.  Following the classic technique, these tools try to find one
linear ranking-function for each loop and combine them multiplicatively
or additively, resulting in a loose quadratic bound or no bound at all.

@@

\ifshort{
\begin{figure*}
\small
\begin{mathpar}
\Rule{Q:Update}
{ q'_{xy}, q'_{yx} \in \Qplusz
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
%
\and \Rule[leftskip=.1cm,rightskip=.1cm]{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\\ \Rule[leftskip=1cm,rightskip=.5cm]{Q:IncP}
{ \Gamma \models y \ge 0 \!
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\} \!
\\\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx} \!
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule[leftskip=.5cm,rightskip=1cm]{Q:Inc}
{ M = M_u + M_e(x {\pm} y)
\\\\ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\\ \Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\ \!\! \Gamma_1 \models \Gamma_2
\\\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ \Gamma'_2 \models \Gamma'_1
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\\ \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy} {\in} \N \, . \, \Gamma \models l_{xy} \le |\inter x y| \}
\\ \mathcal U = \{ xy \mid \exists u_{xy}{\in} \N \, . \, \Gamma \models |\inter x y| \le u_{xy} \}
\\\\ \forall i \in \mathcal U.\, q'_i \ge q_i - r_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U {\cup} \mathcal L {\cup} \{0\} .\, q'_i \ge q_i
\\ q'_0 \geq q_0 {\textstyle + \sum_{i \in \mathcal U} u_i r_i - \sum_{i \in \mathcal L} l_i p_i}
}
{ Q' \succeq_\Gamma Q }

@@

Most interesting are the rules {\sc Q:IncP}\iffull{, {\sc Q:DecP},}{} and {\sc Q:Inc}
which describe how the potential is distributed after a size change of
a variable.
\iffull{The rule {\sc Q:IncP} is for increments $x \gets x + y$
and {\sc Q:DecP} is for decrements $x \gets x - y$, they both apply only
when we can deduce from the logical context $\Gamma$ that $y \ge 0$.
Of course, we have symmetrical rules {\sc Q:IncN} and {\sc Q:DecN}
that can be applied if $y$ is not positive.}{The rule {\sc Q:IncP} is for increments
$x \gets x+y$, it applies when we can deduce from the logical context
$\Gamma$ that $y \ge 0$.  Of course, we have symmetrical rules that
can be applied for decrements and when $y$ is not positive.}
The rules are all equivalent in the case where $y=0$.
Finally, the rule {\sc Q:Inc} can be applied
if we cannot deduce any information about $y$.  In the implementation,
the rules for increments and decrements are combined into one syntax
directed rule where the specialized rules take priority over {\sc Q:Inc}.

\iffull{To explain how rules for increment and decrement work, it is
sufficient to understand the rule {\sc Q:IncP}.  The others follow the
same idea and are symmetrical.}{}  In {\sc Q:IncP}, the program updates a
variable $x$ with $x+y$ where $y \ge 0$.  Since $x$ is changed, the
quantitative annotation must be updated to reflect the change of the
program state.  We write $x'$ for the value of $x$ after the assignment.
Since $x$ is the only variable changed, only intervals of the form
$\inter u x$ and $\inter x u$ will be resized.  Note that for any $u$,
$\inter x u$ will get smaller with the update, and if $x' \in \inter x
u$ we have $|\inter x u| = |\inter x {x'}| + |\inter {x'} u|$.  But
$|\inter x {x'}| = |\inter 0 y|$ which means that the potential
$q'_{0y}$ in the postcondition can be increased by $q_{xu}$ under the
guard that $x' \in \inter x u$.  Dually, the interval $\inter v x$ can
get bigger with the update.  We know that $|\inter v {x'}| \le y +
|\inter v x|$.  So we decrease the potential of $\inter 0 y$ by
$q_{vx}$ to pay for this change.  The rule ensures this
only for $v \not\in \mathcal U$ because we know that $x \le v$
otherwise, and thus $|\inter v x| = 0$.

\iffull{
Another interesting rule is {\sc Q:Call}.  It needs to account for the
changes to the stack caused by the function call, the arguments/return
value passing, and the preservation of local variables.  We can sum up
the main ideas of the rule as follows.
\itemskip
\begin{itemize}
\item The potential in the pre- and postcondition of the function
  specification is equalized to its matching potential in the callee's
  pre- and postcondition.
\itemskipIn
\item
  $\!$The potential of intervals $|\inter x y|$ is preserved
  if $x$ and $y$ are local\iffull{ variables.}{.}
\item
  The unknown potentials after the call (e.g.
  $|\inter x g|$, with $x$ local and $g$ global)
  are set to zero in the postcondition.
\end{itemize}
\itemskip
If $x$ and $y$ are local variables and $f(x,y)$ is
called, {\sc Q:Call} splits the potential of
$|\inter x y|$ in two parts, one part to perform the
computation in the function $f$ and one part to keep for
later use after the function call.  This splitting is
realized by the equations $Q = P {+} S$ and $Q' =
P' {+} S'$.  Arguments in the function precondition
$(\Gamma_f, Q_f)$ are named using a fixed vector $\Vargs$
of names different from all program variables.  This
prevents name conflicts from happening and ensures that
the substitution $[\Vargs/\vec x]$ is meaningful.
Symmetrically, we use the unique name $\Vret$ to represent
the return value in the function's postcondition
$(\Gamma'_f, Q'_f)$.
}{}
